---
# Green_Teams ----
# Written by Chris Snyder, MPH (Roqarr)
# Written for the Montclair State University PSEG Institute for Sustainability Studies
# Documentation ----
# This is the RMD script for powerpoint document generation
# Arguments ----
# Defines the parameters to pass into the Rmd document for knitting.
# Parameters allow you to pass arguments into the Rmd document, similar to a function, to create dynamic reports.
# Similar to a function, you must define the parameters at the beginning of the report, and can assign default values.
# These parameters can be accessed inside the Rmd document using a special environment object: 'params'
# The 'params' object is a special read-only environment object.
# Each of the parameters lsited below, can be accesed as vectors of 'params', using `params$value`
params:
 state: "New Jersey"
 university: "Montclair State University"
 program: "Green Teams"
# Returns ----
# na

# The Title of the Document. Accepts Multiple Lines and Images, when preceeded by the '|', the pipe character.
# By default, this is centered. 
title: | 
   Loading, Saving, and Processing Data
    
 
# The Subtitle of the Document. Accepts Multiple Lines and Images, when preceeded by the '|', the pipe character.
subtitle: |
 How to import, export, and manipulate  
 data inside of R, from local and remote sources

# The List of Authors, separated by a new line and preceeded by a '-' and a space
# Authors will be placed on a line, and automatically centered
# Notes about the authors, as footnotes, can be added with '^[]', with the contents of the footnote placed inside of the brackets.
author: 
  - Chris Snyder, MPH

# Date; Use Pipe to allow multiple lines or inline R code.
# R code can be accepted in most arguments using the backtick / backquote character '`' (located above the tab key).
# This example will use an R function to pull the current date, and add it to the document as the entry for the date.
# You could alternatively write in text for a fixed Year, Semester, Quarter, etc.
date: |
  Green Teams, 
  `r format(Sys.Date(), "%Y")`

# The abstract block of text, located below the authors / date
# abstract: |
#  Here is placeholder abstract Text.
#  It can have multiple lines, and should be limited to a brief description of the paper, if you choose to add one.

# Table of Contents Settings
# Do you want to add a Table of Contents? True or False
toc: True
# The Defined Title of the Table of Contents
# You can specify a title for the Table of Contents here. The example asserts that the toc is called 'Table of Contents'.
toc-title: "Table of Contents"
# The Defined depth of headers to include in the Table of Contents. 
# This value impacts the TOC and Sidebar Navigation in PDF views.
# Provide a number of how many headings / subheadings you want to have presented in the Table of Contents.
# This example will show Headings (1) and Sub-Headings (2), but will not display any section headings at 3 or below.
toc-depth: 2
# Do you want to add a Table of Figures? True or False
# This will add a separate table, listing all of the defined figures. 
# Recommended to keep False
lof: False 
# Do you want to add a Table of Tables? True or False
# This will add a separate table, listing all of the defined tables. 
# Recommended to keep False
lot: False 

# Defines the output file format of the document that will be made by this Rmd file.
# This example provides the three most common document types: PDF, Word, and Web (HTML).
# There are additional potential formats for presentations, powerpoints, etc.
# This example is formatted for a PDF output. 
# Please note, the document format may impact what code or outputs can be accepted inside the Rmd file.
# output: powerpoint_presentation
# The Reference Document is provided
output: 
  powerpoint_presentation:
    reference_doc: Parallax_Blue.pptx

# output: pdf_document
# output: word_document
# output: html_document



header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}

---

# Goals

By the end of the Presentation, we will have learned:

- How to load data from a variety of local formats
- How to load data from a remote source
- How to identify common types of data inside of R
- How to add, remove, sort, and filter the data
- How to save data to a variety of local formats



# What Data Formats can R Use?

- Base R supports reading data from plain text files (.txt) and comma-separated tables (.csv), and compressed RData files
- `readxl` supports Excel (.xls, .xlsx)
- `foreign` supports SAS, SPSS, and dBase (.dbf)
- `RCurl` supports URL, HTTP, FTP
- `jsonlite` supports JSON
- `shapefiles` supports ESRI shapefiles (.shp, .shx, .dbf)

# Getting our Data Table Ready to Use in R

:::::: {.columns}
::: {.column}
Table Headers

- The first row of your data table
- Names should start with a character
- Names must be unique, or will be iterated
- Names should be descriptive or unique IDs
- Names should avoid iterative counting (Q1, Q2, Q3, ...)

:::

::: {.column}
Table Contents

- Every row after the first is read as your data
- Remove any test data, total rows, or metadata before importing
- Remove empty rows, columns, and table formatting
- Any missing cell becomes a value of `NA`

:::
::::::

# What format is our Data?

What file type is our data stored in?

- Identify the file type from the file extension. You may need to turn file extensions on in your file explorer (Windows Explorer, Finder, etc)
- Check the list of installed packages in R, or CRAN, for a package that can read the file format, or install any new packages that can read that file format, if missing
- Load the required package(s) using  `library()`

# Where is our Data Located?

Where are the data saved to? Data can be on local, removable, or networked locations.

- Files should be kept in a permanent location whenever possible
- Ideally, we will use a relative file path, inside the project:  
(`"./my_folder/my_file"`)
- We can also use an absolute file path, within the local file system:  
(`"C:/Users/my_user/my_folder/my_file"`)
- We can use a URL path, for data available on a networked file system:  
(`"https://www.MyWebsite.com/myPage/my_file.csv"`)


# Loading our Data Table into R

Load the required package(s), and read in the data.

- `read.csv("./Data/my_file.csv")`
- `library(readxl)`  
`read_excel("./Data/my_file.xlsx")`
- `load("./Data/my_data.Rdata")`


# Assigning Data to Objects in R

R can store any form of data as an `object` within the R environment.
R objects will most frequently be stored in the `global environment`.
Objects are assigned using the assignment operators: `=` or `<-`. Both operators will generally behave the same.

- `my_csv = read.csv("./Data/my_file.csv")`
- `my_xlsx <- read_excel("./Data/my_file.xlsx")`


# Different Types of Data In R

All `objects` inside of R will be of a specific `class`. Common data `class` are listed below.

- Data Frame:  
A 2-dimensional table, storing multiple types of data
- List:  
A 1-dimensional list, storing multiple types of data
- Vector:  
A 1-dimensional list, storing only data of the same type


# Different Types of Data In R

All `objects` inside of R will be of a specific `type`. Common data `types` are listed below.

- Numeric: `.5`, `1`, `1.5`
- Integer: `1`, `2`, `3`
- Character: `"Anything in Quotes"`, `"1"`
- Logical: `TRUE`, `FALSE`
- Factor: `Strongly Disagree`, `Disagree`, `Agree`, `Strongly Agree`


# Viewing Data in R

Any object can be viewed by referencing it in the R terminal.  
For example: `mtcars` will print the mtcars data frame inside of the console.

The `View()` function is used to display objects in a separate tab. If using `View()` inside of R, you will receive a new pop-up window with your requested object. If using `View()` inside of RStudio, you will receive a new tab within your RStudio session.  
For example: `View(mtcars)` will open the data frame inside of a new tab. We will be using the `mtcars` data set in this presentation.
Note that `View()` is one of the few functions in R that begins with capitalization.

# Data Lab

The rest of the presentation corresponds to the Data Lab.

Launch RStudio and open the lab now if you have not done so already.


# Example Data Set `mtcars`

```{r echo=FALSE, message=FALSE, warning=FALSE}
# This chunk uses some examples from the mtcars dataset. 
# This dataset is pre-loaded into R, and is commonly used as an example data set for statistical tests and graphing.
# However, this data frame has the row names assigned as the make and model of the automobile, as we will see in the Table section below.
# This code will create a new data frame, removing the row names, and will add a vector of those names instead, which we can work with much easier.
kmtcars = mtcars # Make a copy of mtcars to process
kmtcars$model = rownames(mtcars) # Add a vector of data
rownames(kmtcars) = NULL # Set the row names to NULL, which returns it to the default numbering system
kmtcars = select(kmtcars, model, everything()) # Make a new data table, where the model name comes first, then have all other rows in their original order

# Subset the data frame, choosing only the following columns.
# Functional Example
skmtcars = select(kmtcars,
                  model, 
                  mpg,
                  cyl,
                  disp,
                  hp
                  )

# Order the data frame in decreasing order of MPG
# Assign the order of the data frame to be in descending order of MPG. 
# In case of a tie, the first column value is used as a tie-breaker (i.e., is sorted alphabetically).
# skmtcars = skmtcars[order(skmtcars$mpg, decreasing = T),] 
# skmtcars = skmtcars[skmtcars$cyl==4,] # Subset the data frame, to only include cars where cyl == 4; Cars with 4 Cylinder Engines
skmtcars = head(skmtcars, n=5) # Head takes the top 6 (default) rows of a table
# rownames(skmtcars) = NULL # Reset the row numbers, otherwise, they will appear as a new column in our table.

# Draw a pretty table into the rendered output
# kbl, shorthand of kable, renders tables in our output file of choice, in a more readable fashion than simply calling the table
kbl(skmtcars, # The data frame to print
  booktabs = T, # Use the booktabs package, loaded in the YAML section. This is generally recommended.
  escape = F, # This time the escape is F, because we want to process some escape characters; i.e., the '\n' used in our line breaks for our multi-row column names.
  longtable = T, # Use the longtable package, loaded in the YAML section. This supports tables that are long, supports tables that can wrap across pages. It also centers the table on the page. This is generally recommended.
  format = "pipe", # The document format. For this example, it is a powerpoint document, where the table needs to be 'piped' to the renderer, so our format is pipe.
  linesep = "", # A separator to include after each line / a set of lines. The default value puts a space every 5 items. This removes the space.
  col.names = linebreak(c("Make and Model", 
                          "Miles per Gallon",
                          "Number of Cylinders",
                          "Displacement in Cubic Inches",
                          "Gross Horsepower"
                          ),
                        align = c("l", 
                                  "c",
                                  "c",
                                  "c",
                                  "c"
                                  )
                        ),
  align = c("l",
            "c",
            "c",
            "c",
            "c",
            "c"
            ), caption = "Sample of the mtcars data set"
  )


```

# Referencing Different Data in R

How can we reference our data, once is it loaded?

- The object name is used to reference the entire object:  
`mtcars` references the entire mtcars data frame
- Named vectors are used to reference a sub-set of an object:  
`mtcars$mpg` references the vector `mpg`
- Numbered indexes are used to reference a sub-set of a vector:  
`mtcars$mpg[1]` references the first entry of the vector
- Numbered indexes are used to reference a sub-set of an object:  
 `mtcars[1, 2]` references the first row and second column

# Logical Operators

Logical operators are used to sort and filter data in R.

- Less Than: `<`
- Greater Than: `>`
- Equal To: `==`
- Not Equal To: `!=`
- In: `%in%`

# Sorting Data Frames in R, with One Parameter

Data frames can be sorted by their contents, typically by searching for given values within a specified column using logical operators.
The sorting requires at least one specified vector and one desired value.

- To sort a data frame using one desired value in one vector:  
`mtcars[mtcars$am == 1,]`
- To sort a data frame using multiple desired values in one vector:  
`mtcars[mtcars$cyl %in% c(4, 6),]` or  
`mtcars[mtcars$cyl != 8,]`

# Sorting Data Frames in R, with Multiple Parameters

Additional logical operators can be used in this index sorting method to make more complex filters.

- To sort a data frame where either condition may be met:  
`mtcars[mtcars$am == 1 | mtcars$cyl %in% c(4, 6),]`
- To sort a data frame where both conditions must be met:  
`mtcars[mtcars$am == 1 & mtcars$cyl %in% c(4, 6),]`

# Sorted Data Example

```{r echo=FALSE, message=FALSE, warning=FALSE}
kmtcars = mtcars # Make a copy of mtcars to process

kmtcars = kmtcars[kmtcars$am == 1 & kmtcars$cyl %in% c(4, 6),]
# kmtcars = kmtcars[order(kmtcars$mpg, decreasing = T),]

skmtcars = head(kmtcars) # Name Placeholder for Sorting Operations

rownames(skmtcars) = NULL # Reset the row numbers, otherwise, they will appear as a new column in our table.

kbl(skmtcars, # The data frame to print
  booktabs = T, # Use the booktabs package, loaded in the YAML section. This is generally recommended.
  escape = F, # This time the escape is F, because we want to process some escape characters; i.e., the '\n' used in our line breaks for our multi-row column names.
  longtable = T, # Use the longtable package, loaded in the YAML section. This supports tables that are long, supports tables that can wrap across pages. It also centers the table on the page. This is generally recommended.
  format = "pipe", # The document format. For this example, it is a powerpoint document, where the table needs to be 'piped' to the renderer, so our format is pipe.
  linesep = "", # A separator to include after each line / a set of lines. The default value puts a space every 5 items. This removes the space.
  caption = "First Six cars from `mtcars` Data Set, with 4 or 6 cylinders and manual transmission. \n`mtcars[mtcars$am == 1 & mtcars$cyl %in% c(4, 6),]`"
  )


```

# Ordering Data Frames in R

Data frames can be ordered by their contents, by reordering the data based on ascending / descending values of one vector.

- To sort a data frame of highest to lowest values of `mpg`:  
`mtcars[order(mtcars$mpg, decreasing = T),]`

# Sorted and Reordered Data Example

```{r echo=FALSE, message=FALSE, warning=FALSE}
kmtcars = mtcars # Make a copy of mtcars to process

kmtcars = kmtcars[kmtcars$am == 1 & kmtcars$cyl %in% c(4, 6),]
kmtcars = kmtcars[order(kmtcars$mpg, decreasing = T),]

skmtcars = head(kmtcars) # Name Placeholder for Sorting Operations

rownames(skmtcars) = NULL # Reset the row numbers, otherwise, they will appear as a new column in our table.

kbl(skmtcars, # The data frame to print
  booktabs = T, # Use the booktabs package, loaded in the YAML section. This is generally recommended.
  escape = F, # This time the escape is F, because we want to process some escape characters; i.e., the '\n' used in our line breaks for our multi-row column names.
  longtable = T, # Use the longtable package, loaded in the YAML section. This supports tables that are long, supports tables that can wrap across pages. It also centers the table on the page. This is generally recommended.
  format = "pipe", # The document format. For this example, it is a powerpoint document, where the table needs to be 'piped' to the renderer, so our format is pipe.
  linesep = "", # A separator to include after each line / a set of lines. The default value puts a space every 5 items. This removes the space.
  caption = "Top Six cars from `mtcars` Data Set by fuel economy, with 4 or 6 cylinders and manual transmission."
  )

```

# Adding New Vectors to Data

Lists and Data frames store data in `vectors`. A Data frame is practically just a `list` of `vectors`.  

- New vectors can be added to an object by assigning a value to a `vectorized` reference, using the vector operator `$`:  
`mtcars$model = rownames(mtcars)`
- A constant value can be applied to all rows using vectorized reference:  
`mtcars$axle = 2`
- A formula can be applied to all rows using vectorized reference:  
`mtcars$kpl = mtcars$mpg * .354`
<!-- https://www.mpgtokpl.com/conversionformulas.htm -->

# Data with New Vectors Example

```{r echo=FALSE, message=FALSE, warning=FALSE}
kmtcars = mtcars # Make a copy of mtcars to process

kmtcars$kpl = kmtcars$mpg * .354
kmtcars$kpl = round(kmtcars$kpl, 2)
kmtcars$axle = 2

skmtcars = head(kmtcars) # Name Placeholder for Sorting Operations

rownames(skmtcars) = NULL # Reset the row numbers, otherwise, they will appear as a new column in our table.

kbl(skmtcars, # The data frame to print
  booktabs = T, # Use the booktabs package, loaded in the YAML section. This is generally recommended.
  escape = F, # This time the escape is F, because we want to process some escape characters; i.e., the '\n' used in our line breaks for our multi-row column names.
  longtable = T, # Use the longtable package, loaded in the YAML section. This supports tables that are long, supports tables that can wrap across pages. It also centers the table on the page. This is generally recommended.
  format = "pipe", # The document format. For this example, it is a powerpoint document, where the table needs to be 'piped' to the renderer, so our format is pipe.
  linesep = "", # A separator to include after each line / a set of lines. The default value puts a space every 5 items. This removes the space.
  caption = "Top Six cars from `mtcars` Data Set, with Kilometers per Liter and Number of Axles added."
  )

```

# Selecting Specific Vectors of Data

Data frames can be manipulated easier using the `dplyr` package. This package offers support for manipulating data frame objects.

- Data frames can be reduced using the `select` function:  
`library(dplyr)`  
`mtcars = select(mtcars, model, kpl, cyl, hp, axle)`
                  
# Data with Select Vectors Example

```{r echo=FALSE, message=FALSE, warning=FALSE}
kmtcars = mtcars # Make a copy of mtcars to process

kmtcars$kpl = kmtcars$mpg * .354
kmtcars$kpl = round(kmtcars$kpl, 2)
kmtcars$axle = 2
kmtcars$model = rownames(kmtcars)

kmtcars = select(kmtcars, model, kpl, cyl, hp, axle)

skmtcars = head(kmtcars) # Name Placeholder for Sorting Operations

rownames(skmtcars) = NULL # Reset the row numbers, otherwise, they will appear as a new column in our table.

kbl(skmtcars, # The data frame to print
  booktabs = T, # Use the booktabs package, loaded in the YAML section. This is generally recommended.
  escape = F, # This time the escape is F, because we want to process some escape characters; i.e., the '\n' used in our line breaks for our multi-row column names.
  longtable = T, # Use the longtable package, loaded in the YAML section. This supports tables that are long, supports tables that can wrap across pages. It also centers the table on the page. This is generally recommended.
  format = "pipe", # The document format. For this example, it is a powerpoint document, where the table needs to be 'piped' to the renderer, so our format is pipe.
  linesep = "", # A separator to include after each line / a set of lines. The default value puts a space every 5 items. This removes the space.
  caption = "Top Six cars from `mtcars` Data Set, with Kilometers per Liter and Number of Axles added, and other columns removed."
  )

```

# Save our New Data

Data can be written back out from R into files, which can be loaded at a later time, or used in other software. Data can be written to a variety of file types, and does not need to return to its original format. The `save()` function creates a binarized file that only R can read, but is very fast, space efficient, and can have non-normal-shaped data.

- `write.csv(mtcars, file = "./mtcars.csv")`
- `save(mtcars, file = "./mtcars.RData")`

# Goals

At the end of the Presentation, we have learned:

- How to load data from a variety of local formats
- How to load data from a remote source
- How to identify common types of data inside of R
- How to add, remove, sort, and filter the data
- How to save data to a variety of local formats

